---
title: "Phenome-Wide Association Study"
author: "Brian M Schilder"
date: "Vignette updated: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i>"
output:
  rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{monarchr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include=FALSE, message=FALSE}
options(width = 300)
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, fig.width = 10, message = FALSE, warning = FALSE)
```

# Motivation

Phenome-Wide Association Studies (PHEWAS) seek to understand how genetic variants known to cause one disease can have pleiotropic effects on risk for other diseases as well. While this approach has been widely used in the field, gathering the necessary data to conduct such a study comprehensively is a laborious process. Furthermore, the variant-phenotype associations that are required to perform such PHEWAS are frequently changing as new studies are published. Inconsistencies in the level at which the traits are annotated (e.g. cardiovascular disease vs. its subset cardiomyopathy) . Here, we demonstrate how `monarchr` can be used to perform a comprehensive PHEWAS in four steps that can be run in under XX seconds.

## 1. Get all subtypes of a disease

Get cardiovascular disease parent node MONDO ID:

```{r}
library(monarchr)
library(tidygraph)

# Append preferences
monarch <- monarch_engine()
preferences <- monarch$preferences
preferences$category_priority <- c(preferences$category_priority,
																	 "biolink:SequenceVariant")
# Search for cardiovascular disorders
query <- "vascular disorder"
cvd <- monarch_search(query = query, 
											category = "biolink:Disease", 
											preferences = preferences,
											limit=10) 
cvd_subtypes <- cvd|> 
	# activate(nodes) |>
	# filter(grepl(query, name, ignore.case = TRUE)) |>
	expand(predicates = "biolink:subclass_of", 
				 direction = "in",
			 	 transitive = TRUE) 

cvd_subtypes 
```


## 2. Get all genes and variants associated with disease subtypes

```{r}
# monarchr::summarize_neighborhood(cvd_subtypes, direction = "in")

cvd_genetics <- cvd_subtypes |>
	expand(categories = c("biolink:Gene", "biolink:SequenceVariant"), 
				 predicates = c("biolink:causes",
				 							  "biolink:gene_associated_with_condition",
				 							  "biolink:genetically_associated_with"
				 							 )
				 )

table(nodes(cvd_genetics)$pcategory)
```

3. Get other diseases associated with those variants
 
```{r}
# https://biolink.github.io/biolink-model/related_to_at_instance_level/
# "in_linkage_disequilibrium_with"
phewas_hits <- cvd_genetics |> 
	# Get the diseases associated with these variants
	filter(pcategory %in% "biolink:SequenceVariant") |> 
	# Add variants in LD with the original variants
	expand(predicate = "biolink:in_linkage_disequilibrium_with",
				 categories = "biolink:SequenceVariant",
				 direction = "both") |>
	# Get connections from variants to other diseases
	expand(predicates = "biolink:causes",
				 categories = "biolink:Disease", 
				 direction = "out",
				 transitive = TRUE
				 ) |>
	# Get the genes associated with these new variants
	expand(predicates = "biolink:is_sequence_variant_of",
				 categories = "biolink:Gene",
			   direction = "out",
				 transitive = TRUE
				 ) |>
	# Remove the original CVD diseases
	filter(!id %in% nodes(cvd_subtypes)$id) |>
	# Get the subtypes of these diseases
	expand(predicates = "biolink:subclass_of", 
				 direction = "in", 
				 categories = "biolink:Disease", 
				 transitive = TRUE
				 ) 

table(nodes(phewas_hits)$pcategory) 

# Plot the PHEWAS hits
phewas_hits|>
	activate(nodes) |>
	filter(!node_is_isolated()) |>
	plot(layout = "fr")
```

## Merge and plot

Merge the graphs `cvd_genetics`, `phewas_hits`

```{r}
kg_all <- kg_join(cvd_genetics, phewas_hits) |>
	# Add group labels to disease nodes
	mutate(group = ifelse(pcategory == "biolink:Disease", 
													 ifelse(id %in% nodes(cvd_subtypes)$id,
													 			 "Cardiovascular Disorder",
													 			 "Other Disease"), 
													 NA)
						) 
# Count the number of Diseases in each group
nodes(kg_all)|>
	filter(pcategory=="biolink:Disease")|>
	group_by(group)|>count()

# Compute distance from variant and disease nodes
get_node_idx <- function(graph, category){
	which(nodes(graph)$pcategory %in% category)
}
variant_nodes <- get_node_idx(kg_all, "biolink:SequenceVariant")
gene_nodes <- get_node_idx(kg_all, "biolink:Gene")
disease_nodes <- get_node_idx(kg_all, "biolink:Disease")
kg_all <- kg_all|>
	activate(nodes) |>
	mutate(distance_from_variant = node_distance_to(nodes = variant_nodes, mode = "all"),
				 distance_from_gene = node_distance_to(nodes = gene_nodes, mode = "all"),
				 distance_from_disease = node_distance_to(nodes = disease_nodes, mode = "all")) 
# If a variant is associated with a disease, propagate the relationship to the gene with all the descendants of the disease


# Count node distances by category 
nodes(kg_all)|>
	group_by(pcategory)|>
	count(distance_from_gene)|> 
	arrange(pcategory,distance_from_gene)|>
	data.table::data.table() 

table(nodes(kg_all)$pcategory)
table(edges(kg_all)$predicate)
table((edges(kg_all)|> filter(predicate=="biolink:causes"))$primary_knowledge_source)
# table(unlist((edges(kg_all)|> filter(predicate=="biolink:causes"))$provided_by))
table(unlist((edges(kg_all)|> 
								filter(predicate=="biolink:causes"))$qualifiers),useNA = "always")
table(unlist((edges(kg_all)|> 
								filter(predicate=="biolink:causes"))$original_predicate),useNA = "always")


# Weight the edges
# monarchr::kg_edge_weights(kg_all)
rownames(nodes(kg_all)) <- nodes(kg_all)$id

kg_filt <- kg_all |>
	# Filter to only nodes that are somewhat close to the genetic assocation
	# filter(!(pcategory=="biolink:Disease" & distance_from_variant>3) &
	# 			 !(pcategory=="biolink:Gene" & distance_from_variant>3) 
	# 		 	 !(pcategory=="biolink:SequenceVariant" & distance_from_disease>=5)
				 # )|>
	filter(!(pcategory=="biolink:SequenceVariant" & !node_is_adjacent(to = disease_nodes, mode = "all")) &
			  	!(pcategory=="biolink:SequenceVariant" & !node_is_adjacent(to = gene_nodes, mode = "all")) &
				  !(pcategory=="biolink:Gene" & !node_is_adjacent(to = variant_nodes, mode = "all")) &
 			 	  !(pcategory=="biolink:Disease" & distance_from_variant>=4) &
				 	!(pcategory=="biolink:Disease" & distance_from_variant>=4) &
				  !(pcategory=="biolink:Gene" & distance_from_disease>=6)
				 ) |>
	# Drop isolated nodes
	activate(nodes) |>
	filter(!node_is_isolated()) 

table(nodes(kg_filt)$pcategory)
plot(kg_filt, layout = "fr")
	
variant_nodes2 <- get_node_idx(kg_filt, "biolink:SequenceVariant")
# Recompute distance from variant
	kg_filt |>
	mutate(adjacent_to_variant = node_is_adjacent(to = variant_nodes2,
																										 mode = "all")) |>
	# Remove Gene nodes that aren't directly connected with SequenceVariant nodes anymore
	filter(!(pcategory=="biolink:Gene" & adjacent_to_variant==FALSE)
				 ) |>
	# Drop isolated nodes
	activate(nodes) |>
	filter(!node_is_isolated()) |>
	plot(layout="fr")
```


